Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-05-10 analysis/paper_draft2.Rmd
Update paper_draft2.Rmd
update to text

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output: 
  rticles::asa_article:
    citation_package: "default"
    keep_tex: TRUE
    #fig_width: 5
    #fig_height: 5
  #output:
#  pdf_document:
#    latex_engine: xelatex
#    number_sections: true
    #fig_width: 4
    #fig_height: 4
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

Figure 1 depicts the bias detection rate with an 95\% credible interval for the n_strikes = 15 scenario. When $b_{curr}$ is high (close to 3) the coverage rate is high, very close to 1. However, when $b_{curr}$ is lower (close to zero), the bias detection rate drops below 0.8. When there is no bias in the current or historical trials ($b_{curr} = b_{hist} = 0$}), our bias detection rate for 95% CI is 0.05 or less. As we place more weight on the historical strike data (as $a$ increases), and the historical strike data is consistent with current information (high compatibility between current and historical trials, the bias detection rate is close to 1, especially when bias is high (upper right-hand corner of plots). Conversely, as alpha decreases, bias detection is only high when there is high bias in the current trial ($b_{curr} >= 1.5). Increasing the number of historical trails leads to an increase in bias detection especially when current and historical trials are compatible.

@@

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 2), when $b_{curr} = 2$, the coverage rate is lower than with an 80\% credible interval; when $b_{curr}=0$, the coverage rate is higher. The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

@@

Figure 3 depicts coverage rates when using a 95\% credible interval. Here, given $b_{curr} = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to infer bias when there is none. When $b_{curr} = 3$, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b_{curr} = \{2, -2\}$), the coverage rate is only high if we have consistent historical information.

Overall, these simulation results indicate that our model can always accurately detect strong bias. As for moderate bias, detection accuracy increases with more historical strike information. The choice of credible interval ultimately depends on how conservative we want to be when inferring bias.

@@

@@

Finally, in practice, historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for the approach presented here if the missing strike data is missing not at random.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "Eric Barón"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output: 
  rticles::asa_article:
    citation_package: "default"
    keep_tex: TRUE
    #fig_width: 5
    #fig_height: 5
  #output:
#  pdf_document:
#    latex_engine: xelatex
#    number_sections: true
    #fig_width: 4
    #fig_height: 4
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

Figure 1 depicts the bias detection rate with an 95\% credible interval for the n_strikes = 15 scenario. When $b_{curr}$ is high (close to 3) the coverage rate is high, very close to 1. However, when $b_{curr}$ is lower (close to zero), the bias detection rate drops below 0.8. When there is no bias in the current or historical trials ($b_{curr} = b_{hist} = 0$}), our bias detection rate for 95% CI is 0.05 or less. As we place more weight on the historical strike data (as $a$ increases), and the historical strike data is consistent with current information (high compatibility between current and historical trials, the bias detection rate is close to 1, especially when bias is high (upper right-hand corner of plots). Conversely, as alpha decreases, bias detection is only high when there is high bias in the current trial ($b_{curr} >= 1.5). Increasing the number of historical trials leads to an increase in bias detection especially when current and historical trials are compatible.

As we decrease the number of strikes, from 15 to 10 to 6 (Figs. 2 and 3), our ability to detect bias decreases. Bias detection is only high when alpha is high (close to 1), current trial bias is high, there is compatibility between current and historical trials, and the number of historical trials is large.

@@

KEEP/supplement? Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 2), when $b_{curr} = 2$, the coverage rate is lower than with an 80\% credible interval; when $b_{curr}=0$, the coverage rate is higher. The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

@@

Overall, these simulation results indicate that our model can accurately detect strong bias when it is present in both the current and historical trials. increasing alpha strengthens our ability to detect bias when trials are compatible. Increasing the number of historical trials, also facilitates bias detection under agreement between trials. Larger numbers of strikes make bias detection more likely. Under low number of strikes (n_strikes = 6), buas detection is only possible when there is strong bias in the current trial, and alpha is high, and high compatibility.

@@

(update)

@@

(update)
Finally, in practice, historical strike data tends to be incomplete at best, because jury-selection records in past cases are hard to get due to court practices and other barriers [@Grosso2017; @Wright2018]. For example, some jurisdictions do not require recording relevant or detailed enough information about prospective jurors who were or could have been struck (e.g., their race or gender). This forces researchers to turn to other data sources, if any, to find that information. And judges and court staff vary in how often they fully complete or maintain the information they are required to record.  This presents an issue for the approach presented here if the missing strike data is missing not at random.

TODO: Add piece about small sample effect and fact that intial prior will dominate in small sample setting.

future work could focus on initial prior choice, and measuring compatibility between cur and hist to develope, data driven choice for alpha.
